---
title: "Exploratory Data Analysis"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo=TRUE)
```

# Overview

This notebook documents the exploratory data analysis undertaken for this project.

# Loading the Data

Load the tidy data set into memory. If it does not exist, then run the notebook that builds it.

```{r}
tidy_data_filepath <- '../data/tidy.Rds'
stopifnot(file.exists(tidy_data_filepath))

df <- readRDS(tidy_data_filepath)
stopifnot(nrow(df) == 192) # accurate number or responders 
print(df, n=6)
```

# Summarize Data

```{r}
summary(df)
```

```{r}
# Number of unique owners.
length(unique(df$owner_id))
```

# Display Session Info

```{r}
utils::sessionInfo()
```

# Load the Standardized Data

```{r}
df_bite <- readRDS("../data/df_clean.rds")

stopifnot(nrow(df_bite) == 32)
```

# Demographics

## Child Count

Number of children per household.

```{r child-count}
# inspect
summary(df_bite$child_cnt)

# missing
df_bite %>% summarise(n_missing = sum(is.na(child_cnt)))

# summary statistics
df_bite %>% summarise(
  n = sum(!is.na(child_cnt)),
  mean = mean(child_cnt, na.rm = TRUE),
  median = median(child_cnt, na.rm = TRUE),
  sd = sd(child_cnt, na.rm = TRUE), 
  min = min(child_cnt, na.rm = TRUE),
  max = max(child_cnt, na.rm = TRUE))

# summary table
df_bite %>% count(child_cnt) %>% mutate(
  percent = round(n / sum(n) * 100, 1))

# Visuals
## barplot
ggplot(df_bite, aes(x = factor(child_cnt))) +
  geom_bar() + labs(
    title = "Children Per Household (bite occurred)",
    x = "Children",
    y = "n") +
  theme_minimal()
```

## Child Gender

Distribution of gender of child involved in bite occurrence.

```{r child-gender}
# inspect
df_bite %>% count(child_gender_std, sort = TRUE)

# check missing
df_bite %>% summarise(sum(is.na(child_gender_std)))

# summary table
df_bite %>% count(child_gender_std) %>% 
  mutate(percent = round(n / sum(n) * 100, 1))

# visual
## barplot
ggplot(df_bite, aes(x = child_gender_std)) + 
  geom_bar() + labs(
    title = "Child Gender Distribution (bite occurred)",
    x = "Child Gender",
    y = "n") + 
  theme_minimal()
```

## Child Age

Age of child involved, expressed in months and years.

```{r child-age}
# inspect values
summary(df_bite$child_age_years)

# check missing
df_bite %>% summarise(
  n_missing = sum(is.na(child_age_years)))

# summary tables 
## months
df_bite %>% summarise(
  n = sum(!is.na(child_age_months)),
  mean = mean(child_age_months, na.rm = TRUE),
  median = median(child_age_months, na.rm = TRUE),
  sd = sd(child_age_months, na.rm = TRUE),
  min = min(child_age_months, na.rm = TRUE),
  max = max(child_age_months, na.rm = TRUE),
  min = min(child_age_months, na.rm = TRUE))

## years
df_bite %>% summarise(
  n = sum(!is.na(child_age_years)),
  mean = mean(child_age_years, na.rm = TRUE),
  median = median(child_age_years, na.rm = TRUE),
  sd = sd(child_age_years, na.rm = TRUE),
  max = max(child_age_years, na.rm = TRUE),
  min = min(child_age_years, na.rm = TRUE))

# visuals 
## histogram
ggplot(df_bite, aes(x = child_age_years)) +
  geom_histogram(binwidth = 1) + labs(
    title = "Child Age Distribution (bite occurred)",
    x = "Child Age (Years)",
    y = "n") +  
  theme_minimal()

## boxplot
ggplot(df_bite, aes(y = child_age_years)) +
  geom_boxplot() + labs(
    title = "Child Age Boxplot (bite occurred)",
    y = "Child Age (Years)") +
  theme_minimal()
```

## Dog sex

Sex of dog involved in bite: Male/Female categories (regardless of neuter-status).

```{r dog-sex}
# inspect
df_bite %>% count(dog_sex_std, sort = TRUE)

# missing
df_bite %>% summarise(sum(is.na(dog_sex_std)))

# summary tables
df_bite %>% count(dog_sex_std) %>% mutate(
  percent = round(n / sum(n) * 100, 1))

# Visuals
## barplot
ggplot(df_bite, aes(x = dog_sex_std)) +
  geom_bar() +
  labs(title = "Dog Sex Distribution (bite occurred)",
       x = "Sex",
       y = "n") +
  theme_minimal()
```

## Neuter status

Reproductive sterilization (TRUE = neutered/spayed, FALSE = intact).

```{r neuter-status}
# inspect values
df_bite %>% count(dog_neuter, sort = TRUE)

# check missing 
df_bite %>% summarise(sum(is.na(dog_neuter)))

# summary table
df_bite %>% count(dog_neuter) %>% mutate(
  neuter_status = case_when(
    dog_neuter == TRUE ~ "Sterilized",
    dog_neuter == FALSE ~ "Intact",
    TRUE ~ NA_character_),
  percent = round(n / sum(n) * 100, 1)) %>% 
  select(neuter_status, n, percent)

# visual
## barplot
ggplot(df_bite, aes(x = factor(
  dog_neuter,
  levels = c(FALSE, TRUE),
  labels = c("Intact", "Sterilized")))) +
  geom_bar() + labs(
    title = "Dog Sterilization Status (bite occurred)",
    x = "Sterilization Status",
    y = "n") +
  theme_minimal()

## dog sex by neuter status
df_bite %>% filter(
  !is.na(dog_sex_std), !is.na(dog_neuter)) %>% 
  count(dog_sex_std, dog_neuter) %>% 
  ggplot(aes(
    x = dog_sex_std, 
    y = n, 
    fill = factor(dog_neuter, 
                  levels = c(FALSE, TRUE),
                  labels = c("Intact", "Sterilized")))) + 
  geom_col(
    position = "dodge") + labs(
      title = "Dog Sex by Neuter Status (bite occurred)",
      x = "Sex",
      y = "n",
      fill = "Sterilization Status") + 
  theme_minimal()
```

## Dog age

Age of dog involved, expressed in months and years.

```{r dog-age}
# inspect values
summary(df_bite$dog_age_years)

# check missing
df_bite %>% summarise(
  n_missing = sum(is.na(dog_age_years)))

# summary tables 
## months
df_bite %>% summarise(
  n = sum(!is.na(dog_age_months)),
  mean = mean(dog_age_months, na.rm = TRUE),
  median = median(dog_age_months, na.rm = TRUE),
  sd = sd(dog_age_months, na.rm = TRUE),
  min = min(dog_age_months, na.rm = TRUE),
  max = max(dog_age_months, na.rm = TRUE),
  min = min(dog_age_months, na.rm = TRUE))

## years
df_bite %>% summarise(
  n = sum(!is.na(dog_age_years)),
  mean = mean(dog_age_years, na.rm = TRUE),
  median = median(dog_age_years, na.rm = TRUE),
  sd = sd(dog_age_years, na.rm = TRUE),
  max = max(dog_age_years, na.rm = TRUE),
  min = min(dog_age_years, na.rm = TRUE))

# visuals 
## histogram
ggplot(df_bite, aes(x = child_age_years)) +
  geom_histogram(binwidth = 1) + labs(
    title = "Dog Age Distribution (bite occurred)",
    x = "Dog Age (Years)",
    y = "n") +  
  theme_minimal()

## boxplot
ggplot(df_bite, aes(y = child_age_years)) +
  geom_boxplot() + labs(
    title = "Dog Age Boxplot (bite occurred)",
    y = "Dog Age (Years)") +
  theme_minimal()
```

# Summary table

Demographic summaries table with counts and proportions.

```{r demographic-summary}
N <- nrow(df_bite)

bite_demo_table <- tibble(Variable = c(
  "Child gender: Male",
  "Child gender: Female", 
  "Child gender: Prefer not to say",
  "Child age (years)",
  "Child count per household", 
  "Dog sex: Male",
  "Dog sex: Female",
  "Dog sterilized",
  "Dog intact",
  "Dog age (years)"),
  Value = c(
    sprintf("%d (%.1f%%)",
            sum(df_bite$child_gender_std == "Male", na.rm = TRUE),
            100 * sum(df_bite$child_gender_std == "Male", na.rm = TRUE) / N), 
    sprintf("%d (%.1f%%)",
            sum(df_bite$child_gender_std == "Female", na.rm = TRUE),
            100 * sum(df_bite$child_gender_std == "Female", na.rm = TRUE) / N),
    sprintf("%d (%.1f%%)",
            sum(df_bite$child_gender_std == 
                  "Prefer not to say", na.rm = TRUE),
            100 * sum(df_bite$child_gender_std == 
                        "Prefer not to say", na.rm = TRUE) / N),
    sprintf("%.2f (%.2f)",
            mean(df_bite$child_age_years, na.rm = TRUE),
            sd(df_bite$child_age_years, na.rm = TRUE)),
    sprintf("%.2f (%.2f)",
            mean(df_bite$child_cnt, na.rm = TRUE),
            sd(df_bite$child_cnt, na.rm = TRUE)),
    sprintf("%d (%.1f%%)",
            sum(df_bite$dog_sex_std == "Male", na.rm = TRUE),
            100 * sum(df_bite$dog_sex_std == "Male", na.rm = TRUE) / N), 
    sprintf("%d (%.1f%%)",
            sum(df_bite$dog_sex_std == "Female", na.rm = TRUE),
            100 * sum(df_bite$dog_sex_std == "Female", na.rm = TRUE) / N),
    sprintf("%d (%.1f%%)",
            sum(df_bite$dog_neuter == TRUE, na.rm = TRUE),
            100 * sum(df_bite$dog_neuter == TRUE, na.rm = TRUE) / N), 
    sprintf("%d (%.1f%%)",
            sum(df_bite$dog_neuter == FALSE, na.rm = TRUE),
            100 * sum(df_bite$dog_neuter == FALSE, na.rm = TRUE) / N), 
    sprintf("%.2f (%.2f)",
            mean(df_bite$dog_age_years, na.rm = TRUE),
            sd(df_bite$dog_age_years, na.rm = TRUE))))

  
bite_demo_table
```

# Variables

## Dog behavior prior to bite

dog_warning_any, \_count

```{r dog-warning}
df_bite %>% count(dog_warning_any, name = "n") %>% 
  mutate(prop = n / sum(n),
         pct = scales::percent(prop, accuracy = 0.1))

df_bite %>% count(dog_warning_none, name = "n") %>% 
  mutate(prop = n / sum(n),
         pct = scales::percent(prop, accuracy = 0.1))

df_bite %>% count(dog_warning_count, name = "n") %>% 
  mutate(prop = n / sum(n),
         pct = scales::percent(prop, accuracy = 0.1)) %>% 
  arrange(dog_warning_count)

# summary table
df_bite %>% summarise(
  n = n(),
  n_warning_any = sum(
    dog_warning_any == TRUE, na.rm = TRUE),
  pct_warning_any = round(mean(
    dog_warning_any == TRUE, na.rm = TRUE) * 100, 1),
  n_warning_none = sum(
    dog_warning_none == TRUE, na.rm = TRUE),
  pct_warning_none = round(mean(
    dog_warning_none == TRUE, na.rm = TRUE) * 100, 1),
  mean = mean(dog_warning_count, na.rm = TRUE),
  median = median(dog_warning_count, na.rm = TRUE),
  sd = sd(dog_warning_count, na.rm = TRUE),
  min = min(dog_warning_count, na.rm = TRUE),
  max = max(dog_warning_count, na.rm = TRUE))

# visuals
## barplot
ggplot(df_bite, aes(x = factor(
  dog_warning_any,
  levels = c(FALSE, TRUE),
  labels = c("No warning signal", "Warning signal")))) +
  geom_bar() + labs(
    title = "Dog Warning Signal Prior to Bite",
    x = "Warning signal reported",
    y = "n") +
  theme_minimal()

## barplot: number of warning signals
df_bite %>% count(dog_warning_count, name = "n") %>% 
  ggplot(aes(x = factor(dog_warning_count), y = n)) +
  geom_col() + labs(
    title = "Number of Dog Warning Signals Prior to Bite",
    x = "Number of warning signals",
    y = "n") +
  theme_minimal()

## barplot: specific warning signals
dog_warning_cols <- names(df_bite) %>% 
  stringr::str_subset("^dog_behavior_prior_to_bite__") %>% 
  setdiff(c(
    "dog_behavior_prior_to_bite__other",
    "dog_behavior_prior_to_bite__none"))

df_bite %>% 
  summarise(across(all_of(
    dog_warning_cols),
    ~ sum(.x == TRUE, 
          na.rm = TRUE))) %>% 
  pivot_longer(
    everything(),
    names_to = "warning_signal",
    values_to = "n") %>% 
  mutate(
    warning_signal = warning_signal %>% 
      stringr::str_remove("^dog_behavior_prior_to_bite__") %>% 
      stringr::str_replace_all("_", " ") %>% 
      stringr::str_to_sentence()) %>% 
  arrange(n) %>% 
  ggplot(aes(x = reorder(warning_signal, n), y = n)) +
  geom_col() + coord_flip() + labs(
    title = "Dog Warning Signals Reported",
    x = "Warning signal",
    y = "n") +
  theme_minimal()
```

## Dog trigger

dog_trigger_any, \_count

```{r dog-trigger}
# inspect
df_bite %>% count(dog_trigger_any, name = "n") %>% 
  mutate(prop = n / sum(n),
         pct = scales::percent(prop, accuracy = 0.1))

df_bite %>% count(dog_trigger_count, name = "n") %>% 
  mutate(prop = n / sum(n),
         pct = scales::percent(prop, accuracy = 0.1)) %>% 
  arrange(dog_trigger_count)

# summary table
df_bite %>% summarise(
  n = n(),
  n_trigger_any = sum(dog_trigger_any == TRUE, na.rm = TRUE),
  pct_trigger_any = round(mean(dog_trigger_any == TRUE, na.rm = TRUE) * 100, 1),
  mean_trigger_count = mean(dog_trigger_count, na.rm = TRUE),
  median_trigger_count = median(dog_trigger_count, na.rm = TRUE),
  sd_trigger_count = sd(dog_trigger_count, na.rm = TRUE),
  min_trigger_count = min(dog_trigger_count, na.rm = TRUE),
  max_trigger_count = max(dog_trigger_count, na.rm = TRUE))

# visual
## barplot: any trigger
ggplot(df_bite, aes(x = factor(
  dog_trigger_any,
  levels = c(FALSE, TRUE),
  labels = c("No trigger identified", "Trigger identified")))) +
  geom_bar() + labs(
    title = "Bite Triggers Identified",
    x = "Trigger reported",
    y = "n") +
  theme_minimal()

## barplot: number of trigger categories
df_bite %>% count(dog_trigger_count, name = "n") %>% 
  ggplot(aes(x = factor(dog_trigger_count), y = n)) +
  geom_col() +
  labs(
    title = "Number of Bite Trigger Categories",
    x = "Number of trigger categories",
    y = "n") +
  theme_minimal()

## barplot: specific trigger categories
dog_trigger_cols <- names(df_bite) %>% 
  stringr::str_subset("^triggered_bite__") %>% 
  setdiff(c(
    "triggered_bite__other",
    "triggered_bite__other_clean"))

df_bite %>% 
  summarise(across(all_of(
    dog_trigger_cols),
    ~ sum(.x == TRUE, 
          na.rm = TRUE))) %>% 
  pivot_longer(
    everything(),
    names_to = "trigger",
    values_to = "n") %>% 
  mutate(
    trigger = trigger %>% 
      stringr::str_remove("^triggered_bite__") %>% 
      stringr::str_replace_all("_", " ") %>% 
      stringr::str_to_sentence()) %>% 
  arrange(n) %>% 
  ggplot(aes(x = reorder(trigger, n), y = n)) +
  geom_col() + coord_flip() + labs(
    title = "Specific Bite Triggers Reported",
    x = "Trigger",
    y = "n") +
  theme_minimal()
```

## Bite severity

bite_severity

```{r bite-severity}
# inspect 
df_bite %>% count(bite_severity, name = "n") %>% 
  mutate(prop = n / sum(n),
         pct = scales::percent(prop, accuracy = 0.1)) %>% 
  arrange(bite_severity)

# summary table 
df_bite %>% summarise(
  n = sum(!is.na(bite_severity)),
  n_missing = sum(is.na(bite_severity)),
  mean = mean(bite_severity, na.rm = TRUE),
  median = median(bite_severity, na.rm = TRUE),
  min = min(bite_severity, na.rm = TRUE),
  max = max(bite_severity, na.rm = TRUE))

# visual
## barplot
df_bite %>% count(bite_severity, name = "n") %>% 
  ggplot(aes(x = factor(bite_severity), y = n)) +
  geom_col() + labs(
    x = "Bite severity score",
    y = "n",
    title = "Distribution of bite severity")
```

## Dog personality

dog_personality_any, \_count

```{r dog-personality}
# inspect
df_bite %>% count(dog_personality_any, name = "n") %>% 
  mutate(prop = n / sum(n),
         pct = scales::percent(prop, accuracy = 0.1))

df_bite %>% count(dog_personality_count, name = "n") %>% 
  mutate(prop = n / sum(n),
         pct = scales::percent(prop, accuracy = 0.1)) %>% 
  arrange(dog_personality_count)

# summary table
df_bite %>% summarise(
  n = n(),
  n_dog_personality_any = sum(dog_personality_any == TRUE, na.rm = TRUE),
  pct_dog_personality_any = round(mean(
    dog_personality_any == TRUE, na.rm = TRUE) * 100, 1),
  mean_dog_personality_count = mean(
    dog_personality_count, na.rm = TRUE),
  median_dog_personality_count = median(
    dog_personality_count, na.rm = TRUE),
  sd_dog_personality_count = sd(
    dog_personality_count, na.rm = TRUE),
  min_dog_personality_count = min(
    dog_personality_count, na.rm = TRUE),
  max_dog_personality_count = max(
    dog_personality_count, na.rm = TRUE))

# visual
## barplot: any dog personality trait
ggplot(df_bite, aes(x = factor(
  dog_personality_any,
  levels = c(FALSE, TRUE),
  labels = c("No trait endorsed", "Trait endorsed")))) +
  geom_bar() + labs(
    title = "Dog Personality Traits Endorsed",
    x = "Any dog personality trait",
    y = "n") +
  theme_minimal()

## barplot: number of dog personality traits
df_bite %>% count(dog_personality_count, name = "n") %>% 
  ggplot(aes(x = factor(dog_personality_count), y = n)) +
  geom_col() + labs(
    title = "Number of Dog Personality Traits Endorsed",
    x = "Number of dog personality traits",
    y = "n") +
  theme_minimal()

## barplot: specific dog personality traits
dog_personality_cols <- names(df_bite) %>% 
  stringr::str_subset("^dog_personality__") %>% 
  setdiff(c(
    "dog_personality__other",
    "dog_personality__other_clean"))

df_bite %>% 
  summarise(across(
    all_of(dog_personality_cols),
    ~ sum(.x == TRUE, na.rm = TRUE))) %>% 
  pivot_longer(
    everything(),
    names_to = "dog_personality",
    values_to = "n") %>% 
  mutate(
    dog_personality = dog_personality %>% 
      stringr::str_remove("^dog_personality__") %>% 
      stringr::str_replace_all("_", " ") %>% 
      stringr::str_to_sentence()) %>% 
  arrange(n) %>% 
  ggplot(aes(x = reorder(dog_personality, n), y = n)) +
  geom_col() +  coord_flip() + labs(
    title = "Specific Dog Personality Traits Reported",
    x = "Dog personality trait",
    y = "n") +
  theme_minimal()
```

## Child personality

child_personality_any, \_count

```{r child-personality}
# inspect
df_bite %>% count(child_personality_any, name = "n") %>% 
  mutate(prop = n / sum(n),
         pct = scales::percent(prop, accuracy = 0.1))

df_bite %>% count(child_personality_count, name = "n") %>% 
  mutate(prop = n / sum(n),
         pct = scales::percent(prop, accuracy = 0.1)) %>% 
  arrange(child_personality_count)

# summary table
df_bite %>% summarise(
  n = n(),
  n_child_personality_any = sum(
    child_personality_any == TRUE, na.rm = TRUE),
  pct_child_personality_any = round(mean(
    child_personality_any == TRUE, na.rm = TRUE) * 100, 1),
  mean_child_personality_count = mean(
    child_personality_count, na.rm = TRUE),
  median_child_personality_count = median(
    child_personality_count, na.rm = TRUE),
  sd_child_personality_count = sd(
    child_personality_count, na.rm = TRUE),
  min_child_personality_count = min(
    child_personality_count, na.rm = TRUE),
  max_child_personality_count = max(
    child_personality_count, na.rm = TRUE))

# visual
## barplot: any child personality trait
ggplot(df_bite, aes(x = factor(
  child_personality_any,
  levels = c(FALSE, TRUE),
  labels = c("No trait endorsed", "Trait endorsed")))) +
  geom_bar() + labs(
    title = "Child Personality Traits Endorsed",
    x = "Any child personality trait",
    y = "n") +
  theme_minimal()

## barplot: number of child personality traits
df_bite %>% count(child_personality_count, name = "n") %>% 
  ggplot(aes(x = factor(child_personality_count), y = n)) +
  geom_col() + labs(
    title = "Number of Child Personality Traits Endorsed",
    x = "Number of child personality traits",
    y = "n") +
  theme_minimal()

## barplot: specific child personality traits
child_personality_cols <- names(df_bite) %>% 
  stringr::str_subset("^personality_at_bite__") %>% 
  setdiff(c(
    "personality_at_bite__other",
    "personality_at_bite__other_clean"))

df_bite %>% 
  summarise(across(
    all_of(child_personality_cols),
    ~ sum(.x == TRUE, na.rm = TRUE))) %>% 
  pivot_longer(
    everything(),
    names_to = "child_personality",
    values_to = "n") %>% 
  mutate(
    child_personality = child_personality %>% 
      stringr::str_remove("^personality_at_bite__") %>% 
      stringr::str_replace_all("_", " ") %>% 
      stringr::str_to_sentence()) %>% 
  arrange(n) %>% 
  ggplot(aes(x = reorder(child_personality, n), y = n)) +
  geom_col() + coord_flip() + labs(
    title = "Specific Child Personality Traits Reported",
    x = "Child personality trait",
    y = "n") +
  theme_minimal()
```

## Approach direction

approach_direction

```{r approach-direction}
df_bite <- df_bite %>% mutate(
  approach_direction = as.character(approach_direction),
  approach_direction = case_when(
    approach_direction == "child_approached_dog" ~ "child_approached_dog",
    approach_direction == "dog_approached_child" ~ "dog_approached_child",
    approach_direction == "mutual_approach" ~ "mutual_approach",
    approach_direction == "unknown" ~ "unknown",
    TRUE ~ NA_character_),
  approach_direction = factor(
    approach_direction, levels = c(
      "child_approached_dog",
      "dog_approached_child",
      "mutual_approach",
      "unknown")))

# inspect
df_bite %>% count(approach_direction, name = "n") %>% 
  mutate(prop = n / sum(n),
         pct = scales::percent(prop, accuracy = 0.1))

# visual
## barplot: approach direction
ggplot(df_bite, aes(x = approach_direction)) +
  geom_bar() +
  scale_x_discrete(
    drop = FALSE,
    labels = c(
      child_approached_dog = "Child approached dog",
      dog_approached_child = "Dog approached child",
      mutual_approach = "Mutual approach",
      unknown = "Unknown")) + labs(
        title = "Approach Direction Prior to Bite",
        x = "Approach direction",
        y = "n") +
  theme_minimal()

## stacked barplot: bite severity by approach direction
df_bite %>% 
  filter(!is.na(approach_direction), !is.na(bite_severity)) %>% 
  ggplot(aes(
    x = approach_direction,
    fill = factor(bite_severity))) + 
  geom_bar(position = "stack") +
  scale_x_discrete(
    drop = FALSE,
    labels = c(
      child_approached_dog = "Child approached dog",
      dog_approached_child = "Dog approached child",
      mutual_approach = "Mutual approach",
      unknown = "Unknown")) + labs(
        title = "Bite Severity by Approach Direction",
        x = "Approach direction",
        y = "n",
        fill = "Severity") + 
  theme_minimal()
```

# Associations

## warning signs and bite severity

```{r warning-signals}
df_bite %>% count(dog_warning_any, bite_severity_clean)

# visual
ggplot(df_bite, aes(
  x = factor(bite_severity_clean),
  fill = dog_warning_any)) + 
  geom_bar(position = "stack") + 
  scale_fill_discrete(labels = c(
    "FALSE" = "No warning signal",
    "TRUE" = "Warning signal")) +
  labs(
    title = "Warning Signals by Bite Severity",
    x = "Bite Severity",
    y = "Number of Warning Signals") +
  theme_minimal()

# warning count trend by severity
df_bite %>% group_by(bite_severity_clean) %>% 
  summarise(
    warning_count_mean = mean(dog_warning_count, na.rm = TRUE),
    warning_count_median = median(dog_warning_count, na.rm = TRUE),
    n = n())

# warning count distribution 
ggplot(df_bite, aes(
  x = factor(bite_severity_clean),
  fill = factor(dog_warning_count))) + 
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) + 
  labs(
    title = "Warning Signal Distribution Across Severity",
    x = "Bite Severity",
    y = "Proportion of Cases",
    fill = "Warning Signal Count") + 
  theme_minimal()

# Jitter plot
ggplot(df_bite,
       aes(
         x = factor(bite_severity_clean),
         y = dog_warning_count)) +
  geom_jitter(
    width = 0.15,
    height = 0.05,
    alpha = 0.7,
    size = 3) +
  labs(
    title = "Warning Signal Across Bite Severities",
    x = "Bite Severity",
    y = "Warning Signal Count") +
  theme_minimal()
```

## approach direction x warning signals

```{r}
df_bite %>% count(approach_direction, dog_warning_any)

# visual 
ggplot(df_bite, aes(
  x = approach_direction,
  fill = dog_warning_any)) +
  geom_bar(position = "fill") + 
  scale_y_continuous(labels = scales::percent) +
  scale_fill_discrete(labels = c(
    "FALSE" = "No warning signal",
    "TRUE" = "Warning Signal")) +
  scale_x_discrete(labels = c(
    child_approached_dog = "Child approached dog",
    dog_approached_child = "Dog approached child",
    mutual_approach = "Mutual approach",
    unknown = "Unknown")) +
  labs(
    title = "Warning Signals by Approach Direction",
    x = "Approach Direction",
    y = "Proportion of Bite Cases",
    fill = "Warning Signal") + 
  theme_minimal()

# proportionality visual

# define warning rate
warning_summary <- df_bite %>%
  group_by(approach_direction) %>%
  summarise(
    warning_rate = mean(dog_warning_any, na.rm = TRUE),
    n = n())

# visual
ggplot(warning_summary, aes(
  x = approach_direction,
  y = warning_rate)) +
  geom_point(size = 4) +
  geom_segment(aes(
    x = approach_direction,
    xend = approach_direction,
    y = 0,
    yend = warning_rate)) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)) +
  scale_x_discrete(labels = c(
    child_approached_dog = "Child approached dog",
    dog_approached_child = "Dog approached child",
    mutual_approach = "Mutual approach",
    unknown = "Unknown")) +
  labs(
    title = "Proportion of Bite Incidents With Warning Signals",
    x = "Interaction Direction",
    y = "Warning Signal Prevalence") +
  theme_minimal()
```

## child age x severity

```{r}
df_bite %>% group_by(bite_severity_clean) %>% 
  summarise(
    child_age_mean = mean(child_age_years, na.rm = TRUE),
    child_age_median = median(child_age_years, na.rm = TRUE), 
    child_age_sd = sd(child_age_years, na.rm = TRUE),
    min_child_age = min(child_age_years, na.rm = TRUE), 
    n = n(), 
    .groups = "drop")

# visual 
ggplot(df_bite, aes(
  x = factor(bite_severity_clean),
  y = child_age_years)) + 
  geom_boxplot() + geom_jitter(
    width = 0.15,
    alpha = 0.7,
    size = 2) + 
  labs(
    title = "Child Age Across Bite Severity",
    x = "Bite Severity",
    y = "Child Age (Years)") + 
  theme_minimal()
```

## child age x warning signals

```{r}
# summary table
df_bite %>%
  group_by(dog_warning_any) %>%
  summarise(
    mean_child_age = mean(child_age_years, na.rm = TRUE),
    median_child_age = median(child_age_years, na.rm = TRUE),
    sd_child_age = sd(child_age_years, na.rm = TRUE),
    min_child_age = min(child_age_years, na.rm = TRUE),
    max_child_age = max(child_age_years, na.rm = TRUE),
    n = n(),
    .groups = "drop")

# visualization
ggplot(df_bite,
       aes(
         x = dog_warning_any,
         y = child_age_years)) +
  geom_boxplot() +
  geom_jitter(
    width = 0.15,
    alpha = 0.7,
    size = 2) +
  scale_x_discrete(labels = c(
    "FALSE" = "No warning signal",
    "TRUE" = "Warning signal")) +
  labs(
    title = "Child Age by Warning Signal Presence",
    x = "Warning Signal Presence",
    y = "Child Age (Years)") +
  theme_minimal()
```

## dog age x severity

```{r}
# summary table
df_bite %>%
  group_by(bite_severity_clean) %>%summarise(
    mean_dog_age = mean(dog_age_years, na.rm = TRUE),
    median_dog_age = median(dog_age_years, na.rm = TRUE),
    sd_dog_age = sd(dog_age_years, na.rm = TRUE),
    min_dog_age = min(dog_age_years, na.rm = TRUE),
    max_dog_age = max(dog_age_years, na.rm = TRUE),
    n = n(),
    .groups = "drop")

# visualization
ggplot(df_bite, aes(
         x = factor(bite_severity_clean),
         y = dog_age_years)) +
  geom_boxplot() +
  geom_jitter(
    width = 0.15,
    alpha = 0.7,
    size = 2) +
  labs(
    title = "Dog Age Across Bite Severities",
    x = "Bite Severity",
    y = "Dog Age (Years)") +
  theme_minimal()
```

## approach direction x warning signals

```{r dog-age-x-warning}
## dog age x warning signs 
# summary table
df_bite %>%
  group_by(dog_warning_any) %>% summarise(
    mean_dog_age = mean(dog_age_years, na.rm = TRUE),
    median_dog_age = median(dog_age_years, na.rm = TRUE),
    sd_dog_age = sd(dog_age_years, na.rm = TRUE),
    min_dog_age = min(dog_age_years, na.rm = TRUE),
    max_dog_age = max(dog_age_years, na.rm = TRUE),
    n = n(),
    .groups = "drop")

# visualization
ggplot(df_bite, aes(
         x = dog_warning_any,
         y = dog_age_years)) +
  geom_boxplot() +
  geom_jitter(
    width = 0.15,
    alpha = 0.7,
    size = 2) +
  scale_x_discrete(labels = c(
    "FALSE" = "No warning signal",
    "TRUE" = "Warning signal")) +
  labs(
    title = "Dog Age by Warning Signal Presence",
    x = "Warning Signal Presence",
    y = "Dog Age (Years)") +
  theme_minimal()
```

## Dog sex x severity

```{r dog-sex-x-severity}
## dog sex x severity
# counts
df_bite %>% count(dog_sex_std, bite_severity_clean)

# visualization
ggplot(df_bite, aes(
         x = dog_sex_std,
         fill = factor(bite_severity_clean))) +
  geom_bar(position = "stack") +
  scale_x_discrete(labels = c(
    Female = "Female",
    Male = "Male")) +
  labs(
    title = "Bite Severity by Dog Sex",
    x = "Dog Sex",
    y = "Count",
    fill = "Severity") +
  theme_minimal()


```

## Dog sex x warning signal

```{r dog-sex-signal}
## dog sex x signals
# counts
df_bite %>% count(dog_sex_std, dog_warning_any)

# proportional visualization
ggplot(df_bite, aes(
         x = dog_sex_std,
         fill = dog_warning_any)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_discrete(labels = c(
    "FALSE" = "No warning signal",
    "TRUE" = "Warning signal")) +
  scale_x_discrete(labels = c(
    Female = "Female",
    Male = "Male")) +
  labs(
    title = "Warning Signal Presence by Dog Sex",
    x = "Dog Sex",
    y = "Proportion of Cases",
    fill = "Warning Signal") +
  theme_minimal()
```
