---
title: "Building Tidy Data"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo=TRUE)
```

# Overview

This notebook serves as reference for the tidying procedures used to prepare the study data for analysis.

# Form the Data Frame

## Load the Raw Data

The raw data is loaded from a CSV file exported from Typeform.

```{r}
# Ignore the first two columns; they are junk.
df <- readr::read_csv('../data/dogs-and-children-survey.csv')

# Assert that the data set is 192 rows and 676 cols.
stopifnot(identical(dim(df)+0, c(192, 676)))
```

We update the data frame column names to be more friendly to work with during analysis.

```{r}
# Apply analysis friendly column names.
colnames(df) <- readr::read_csv('../data/column-names.csv') %>%
  dplyr::pull('readable')

# Peek at the data.
print(df, n=6)
```

# Drop Unnecessary Columns

We drop columns that will not be used in analysis.

```{r}
df <- df %>%
  select(-c(idx, id, start_date_utc, stage_date_utc, submit_date_utc, tags,
            ending))

# Assert size after dropping columns.
stopifnot(identical(dim(df)+0, c(192, 669)))
```

# Save the Tidy Data

Save the data to a file in the data directory using RDS format so that the data\
types are stored and the resulting file is compressed.

```{r}
saveRDS(df, '../data/tidy.Rds')   
```

# Display Session Info

```{r}
sessionInfo()
```

# Demographic Standardization

Basic demographic summaries and standardization for analysis.

## Load packages

Load packages used in tidy notebook.

```{r packages}
library(tidyverse)
```

## Load data

Load the tidy data from previous session.

```{r load-tidy-data}
tidy_path <- "../data/tidy.Rds"
stopifnot(file.exists(tidy_path))
df <- readRDS(tidy_path)
stopifnot(nrow(df) == 192)
```

## Household identifier

Create household id to identify multiple submissions.

```{r "household-id"}
if (!"household_id" %in% names(df)) {
  df <- df %>% 
    mutate(household_id = 
             dplyr::coalesce(network_id, owner_id) %>% 
             stringr::str_to_lower() %>% 
             stringr::str_trim())}

# total number of responses
nrow(df)

# submissions per house
df %>%  
  count(household_id, name = "n_submissions") %>%
  pull(n_submissions) %>% 
  summary()

# number of households with multiple submissions
df %>%  
  count(household_id, name = "n_submissions") %>%
  summarise(n_multi_submit = sum(n_submissions > 1))

# number of responses reporting at least one bite
df %>% summarise(n_child_bite = sum(
  has_child_been_bitten == 1, na.rm = TRUE))

# bite-exists household
df %>% group_by(household_id) %>% summarise(
  bite_present = any(
    has_child_been_bitten == 1, na.rm = TRUE)) %>% 
  summarise(n_household_bite = sum(bite_present))
```

## Bite flag

Create bite flag object.

```{r bite-flag}
df <- df %>% mutate(
    incident_present = 
      !is.na(child_id) |
      !is.na(child_age_at_bite) |
      !is.na(desc_action_prior_to_bite) |
      if_any(starts_with("bite_location__"),
             ~ !is.na(.x)  & stringr::str_squish(
               as.character(.x)) != "")) %>%
  mutate(bite_flag = case_when(
    has_child_been_bitten == 1 ~ TRUE,
    has_child_been_bitten == 0 ~ FALSE,
    incident_present ~ TRUE, 
    TRUE ~ NA),
    bite_flag_correct = case_when(
      child_cnt == 0 ~ NA, # households without children
      !is.na(bite_flag) ~ bite_flag,
      TRUE ~ FALSE))
  
# checks

df %>% count(has_child_been_bitten, bite_flag, bite_flag_correct)

df %>% summarise(n_total = n(),
                 n_bite_flag_correct = sum(bite_flag_correct, na.rm = TRUE), 
                 n_incident_present = sum(incident_present, na.rm = TRUE),
                 n_children_bitten = sum(has_child_been_bitten == 1, na.rm = TRUE))
```

## Check duplicate submissions

Create household groups to separate multiple submissions: one row per household.

```{r household-summary}
household_summary <- df %>%  
  group_by(household_id) %>% 
  summarise(n_household_submissions = n(),
            n_bite_rows = sum(bite_flag_correct, na.rm = TRUE),
            .groups = "drop")

summary(household_summary$n_household_submissions)

household_summary %>% count(n_household_submissions, sort = TRUE)

household_summary %>% summarise(
  n_duplicate_households = sum(n_household_submissions > 1))
```

## Create "df_tidy"

Remove duplicate submissions

```{r remove-duplicates}
df_tidy <- df %>% 
  arrange(household_id) %>% 
  group_by(household_id) %>% 
  slice(1) %>% 
  ungroup()

# check 
nrow(df)
nrow(df_tidy)

df_tidy %>% 
  count(household_id, name = "n_household_submissions") %>% 
  summarise(max_household_submissions = max(n_household_submissions))
```

# Create dataframes for analysis

## Child count

Character to numeric

```{r child-count}
df_tidy <- df_tidy %>% 
  mutate(child_cnt_raw = as.character(child_cnt), 
         child_cnt_raw = str_to_lower(str_trim(child_cnt_raw)),
         child_cnt = case_when(
           child_cnt_raw %in% c("", "na", "n/a") ~ NA_real_,
           child_cnt_raw %in% c("none", "zero", "0") ~ 0,
           child_cnt_raw %in% c("one", "1") ~ 1,
           child_cnt_raw %in% c("two", "2") ~ 2,
           child_cnt_raw %in% c("three", "3") ~ 3,
           child_cnt_raw %in% c("four", "4") ~ 4,
           child_cnt_raw %in% c("five", "5") ~ 5,
           TRUE ~suppressWarnings(as.numeric(child_cnt_raw))))

# check
df_tidy %>% count(child_cnt, sort = TRUE)

# confirm non-child rows are empty
df_tidy %>% filter(child_cnt == 0) %>% 
  summarise(n_household_submissions = n(),
            n_child_age = sum(!is.na(child_age_at_bite)),
            n_child_id = sum(!is.na(child_id)),
            n_dog_sex = sum(!is.na(dog_sex)))
```

## Non-child rows

Save non-child and child rows and create dataframe to eliminate missingness

```{r no-children}
df_no_child <- df_tidy %>% filter(child_cnt == 0)

df_child <- df_tidy %>% filter(child_cnt > 0)

# check
nrow(df_tidy)
nrow(df_no_child)
nrow(df_child)

df_child %>% count(child_cnt, sort = TRUE) # now we use "df_child"
```

## Bite-only dataframe

Create a dataframe for all bite incidents

```{r df-bites}
df_bite <- df_child %>% 
  filter(bite_flag_correct == TRUE)

# check
nrow(df_child)
nrow(df_bite)

df_bite %>% 
  count(bite_flag_correct, sort = TRUE)
```

# Demographic Standardization

Standardize child count, bite flag, child and dog age, child gender, and create a neuter status variable.

## Child age

Convert to months and years.

```{r child-age}
# see all raw ages
df_bite %>% count(child_age_at_bite, sort = TRUE) %>% 
  print(n = 100)

df_bite <- df_bite %>% 
  mutate(
    child_age_raw = str_to_lower(str_squish(as.character(child_age_at_bite))),
    child_age_num = readr::parse_number(child_age_raw),
    child_age_months = case_when(
      is.na(child_age_raw) | child_age_raw == "" ~ NA_real_,
      str_detect(child_age_raw, "month|months|mo\\b") ~ child_age_num,
      str_detect(child_age_raw, "year|years|yr\\b|yo\\b") ~ child_age_num * 12,
      str_detect(child_age_raw, "^around\\s*[0-9]") ~ child_age_num * 12,
      child_age_raw %in% as.character(0:18) ~ child_age_num *12,
      str_detect(child_age_raw, "^[0-9]+\\.[0-9]+$") ~ child_age_num * 12,
                              TRUE ~ NA_real_),
    child_age_years = child_age_months / 12)

# check
df_bite %>%
  count(child_age_at_bite, child_age_months, sort = TRUE)

# drop columns
df_bite <- df_bite %>% select(-child_age_raw, -child_age_num)

# summary
df_bite %>% select(child_age_months, child_age_years) %>% summary()

class(df_bite$child_age_months) 
class(df_bite$child_age_years) # both numeric
```

## Standardize child gender

Standardize character values.

```{r child-gender}
df_bite <- df_bite %>% mutate(child_sex_raw = str_to_lower(str_squish(child_sex)),
                            child_gender_std = case_when(is.na(child_sex_raw) | child_sex_raw == "" ~ NA_character_,
                                                               child_sex_raw %in% c("female", "f", "girl", "woman") ~ "Female",
                                                               child_sex_raw %in% c("male", "boy", "man") ~ "Male",
                                                               str_detect(child_sex_raw, "non[- ]?binary|nb\\b|genderqueer|trans|other") ~ "Other/Nonbinary",
                                                               str_detect(child_sex_raw, "prefer not|prefer not to say|decline") ~ "Prefer not to say",
                                                               TRUE ~ "Needs review"))
# check
df_bite %>% 
  count(child_gender_std, sort = TRUE)

# verify Needs review
df_bite %>% 
  filter(child_gender_std == "Needs review") %>% 
  count(child_sex_raw, sort = TRUE)

# remove columns
df_bite <- df_bite %>%  
  select(-child_sex_raw)

class(df_bite$child_gender_std) # character
```

## Dog age

Convert to months and years.

```{r dog-age}
# inspect raw values
df_bite %>% count(dog_age_at_bite, sort = TRUE) %>% 
  print(n = 100)

df_bite <- df_bite %>% 
  mutate(
    dog_age_raw = str_to_lower(str_squish(as.character(dog_age_at_bite))),
    dog_age_num = readr::parse_number(dog_age_raw),
    dog_age_months = case_when(
      is.na(dog_age_raw) | dog_age_raw == "" ~ NA_real_,
      str_detect(dog_age_raw, "month|months|mo\\b") ~ dog_age_num, # months
      str_detect(dog_age_raw, "year|years|yr\\b|yo\\b") ~ dog_age_num * 12, # years
      str_detect(dog_age_raw, "^around\\s*[0-9]") ~ dog_age_num * 12, 
      dog_age_raw %in% as.character(0:30)~ dog_age_num *12,
      str_detect(dog_age_raw, "^[0-9]+\\.[0-9]+$") ~ dog_age_num * 12,
      TRUE ~ NA_real_),
    dog_age_years = dog_age_months / 12)

# check
df_bite %>%
  count(dog_age_at_bite, dog_age_months, sort = TRUE)

# drop cols
df_bite <- df_bite %>% 
  select(-dog_age_raw, - dog_age_num)

# summary
df_bite %>% select(dog_age_months, dog_age_years) %>% 
  summary()

class(df_bite$dog_age_months) 
class(df_bite$dog_age_years) # numeric
```

## Dog sex/neuter status

Inspect dog gender to formalize dog_neuter status variable.

```{r neuter-status}
# inspect raw values 
df_bite %>% 
  count(dog_sex, sort = TRUE)

# create variable for sterilization
df_bite <- df_bite %>% 
  mutate(
    dog_sex_raw = str_to_lower(str_squish(dog_sex)),
    dog_sex_std = case_when(
      is.na(dog_sex_raw) | dog_sex_raw == "" ~ NA_character_,
      str_detect(dog_sex_raw, "\\bfemale\\b") ~ "Female",
      str_detect(dog_sex_raw, "\\bmale\\b") ~ "Male",
      TRUE ~ "Needs review"),
    dog_neuter = case_when(
      is.na(dog_sex_raw) | dog_sex_raw == "" ~ NA,
      str_detect(dog_sex_raw, "spayed|neutered") ~ TRUE,
      dog_sex_raw %in% 
        c("male", "female") ~ FALSE, TRUE ~ NA))
                                           
# check
df_bite %>%  count(dog_sex, dog_sex_std, dog_neuter, sort = TRUE)

df_bite %>% 
  filter(dog_sex_std == "Needs review") %>% 
  count(dog_sex_raw, sort = TRUE)

class(df_bite$dog_neuter) # logical

# drop cols
df_bite <- df_bite %>% 
  select(-dog_sex_raw)
```

# Column standardization

Identify and covert columns that need standardization

## Checkbox columns standardization

Identify and save survey items with multiple checkbox items.

```{r checkbox-cols}

# All checkbox items
checkbox_cols <- names(df_bite)[str_detect(names(df_bite), "__")]

# "Other" write-in items
checkbox_other <- checkbox_cols[str_detect(
  checkbox_cols,
  "__other($|_)|other_text|specify|describe")]

# save and convert selected checkbox columns to logical
checkbox_logical <- setdiff(checkbox_cols, checkbox_other)

# check
checkbox_cols
checkbox_other
checkbox_logical
```

## Convert checkbox items

### Convert only true checkbox selections to logical (retain other)

```{r checkbox-conversion}
# unchecked = FALSE; checked = TRUE
df_bite <- df_bite %>%
  mutate(across(all_of(checkbox_logical), ~{
    x <- .x
    if (is.logical(x)) { 
      replace(x, is.na (x), FALSE)
      } else if (is.numeric(x) || is.integer(x)) {
        case_when(
          x == 1 ~ TRUE,
          x == 0 ~ FALSE,
          is.na(x) ~ FALSE,
          TRUE ~ FALSE)
        } else if (is.character(x)) {
          x2 <- str_squish(as.character(x))
          case_when(
            is.na(x2) | x2 == "" ~ FALSE,
            TRUE ~ TRUE )
          } else { 
            FALSE }})) # all others

# retain write-in "other" response as character string
df_bite <- df_bite %>% 
  mutate(across(all_of(checkbox_other), ~{
    x <- str_squish(as.character(.x))
    na_if(x, "")
  }))

# check
table(sapply(df_bite[checkbox_logical], class))
table(sapply(df_bite[checkbox_other], class))

df_bite %>% count(triggered_bite__hugged_the_dog, sort = TRUE)
```

1.  dog_consequence (T/F and character \# Misc column standardization

## Identify columns that need standardization

```{r identify-variables}
df_bite %>% select(where(~ !is.logical(.))) %>% names()

# manually rule out columns that should stay non-logical
non_logical_cols <- c(
  "child_cnt",
  "child_id",
  "child_sex",
  "child_age_at_bite",
  "desc_action_prior_to_bite",
  "bite_severity",
  "dog_age_at_bite",
  "dog_sex",
  "dog_breed_category",
  "dog_purebred_breed",
  "dog_consequences",
  "child_age_at_second_bite",
  "desc_action_prior_to_second_bite",
  "second_bite_severity",
  "desc_additional_bite_incidents",
  "child_age_at_alt_dog_bite",
  "desc_action_prior_to_alt_dog_bite",
  "alt_dog_bite_severity",
  "alt_dog_gender",
  "alt_dog_age",
  "alt_dog_breed_description",
  "alt_dog_consequences",
  "child_age_at_second_alt_dog_bite",
  "desc_action_prior_to_second_alt_dog_bite",
  "second_alt_dog_bite_severity",
  "did_child_comfort_level_change_after_bite",
  "anything_else_to_share",
  "owner_id",
  "location",
  "response_type",
  "network_id",
  "household_id",
  "child_cnt_raw",
  "child_age_months",
  "child_age_years",
  "child_gender_std",
  "dog_age_months",
  "dog_age_years",
  "dog_sex_std",
  checkbox_other) %>% unique()

setdiff(non_logical_cols, names(df_bite))

# convert all others to logical; preserve NA
df_bite <- df_bite %>% 
  mutate(across(-all_of(non_logical_cols), ~ case_when(
    is.na(.) ~ NA,
    . %in% c(1, TRUE) ~ TRUE, 
    . %in% c(0, FALSE) ~ FALSE, 
    TRUE ~ NA)))

# re-check variables
df_bite %>% 
  select(where(~ !is.logical(.))) %>% 
  names()

# confirm retained "other" character strings
table(sapply(df_bite[checkbox_other], class))

# preview "other" responses
df_bite %>% 
  filter(if_any(all_of(checkbox_other), ~ !is.na(.))) %>% 
  select(all_of(checkbox_other)) %>% 
  print(width = Inf)
```

# Save CSV

save "df_bite" as a CSV for future analysis

```{r}
readr::write_csv(df_bite, "bite-data-std.csv")

# confirm save
list.files()
```
