---
title: "Building Tidy Data"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo=TRUE)
```

# Overview

This notebook serves as reference for the tidying procedures used to prepare the study data for analysis.

# Form the Data Frame

## Load the Raw Data

The raw data is loaded from a CSV file exported from Typeform.

```{r}
# Ignore the first two columns; they are junk.
df <- readr::read_csv('../data/dogs-and-children-survey.csv')

# Assert that the data set is 192 rows and 676 cols.
stopifnot(identical(dim(df)+0, c(192, 676)))
```

We update the data frame column names to be more friendly to work with during analysis.

```{r}
# Apply analysis friendly column names.
colnames(df) <- readr::read_csv('../data/column-names.csv') %>%
  dplyr::pull('readable')

# Peek at the data.
print(df, n=6)
```

# Drop Unnecessary Columns

We drop columns that will not be used in analysis.

```{r}
df <- df %>%
  select(-c(idx, id, start_date_utc, stage_date_utc, submit_date_utc, tags,
            ending))

# Assert size after dropping columns.
stopifnot(identical(dim(df)+0, c(192, 669)))
```

# Save the Tidy Data

Save the data to a file in the data directory using RDS format so that the data\
types are stored and the resulting file is compressed.

```{r}
saveRDS(df, '../data/tidy.Rds')   

```

# Display Session Info

```{r}
sessionInfo()
```

# Load Packages

```{r packages}
library(tidyverse)
```

# Load data

Load the tidy data from previous session.

```{r load-tidy-data}
df <- readRDS("../data/tidy.Rds")
stopifnot(nrow(df) == 192)
```

## Household identifier

Create household id to identify multiple submissions.

```{r "household-id"}
if (!"household_id" %in% names(df)) {
  df <- df %>% 
    mutate(household_id = 
             dplyr::coalesce(network_id, owner_id) %>% 
             stringr::str_to_lower() %>% 
             stringr::str_trim())}

# total number of responses
nrow(df)

# submissions per house
df %>%  
  count(household_id, name = "n_submissions") %>%
  pull(n_submissions) %>% 
  summary()

# number of households with multiple submissions
df %>%  
  count(household_id, name = "n_submissions") %>%
  summarise(n_multi_submit = sum(n_submissions > 1))

# number of responses reporting at least one bite
df %>% summarise(n_child_bite = sum(
  has_child_been_bitten == 1, na.rm = TRUE))

# bite-exists household
df %>% group_by(household_id) %>% summarise(
  bite_present = any(
    has_child_been_bitten == 1, na.rm = TRUE)) %>% 
  summarise(n_household_bite = sum(bite_present))
```

## Bite flag

Create bite flag object.

```{r bite-flag}
df <- df %>% mutate(
    incident_present = 
      !is.na(child_id) |
      !is.na(child_age_at_bite) |
      !is.na(desc_action_prior_to_bite) |
      if_any(starts_with("bite_location__"),
             ~ !is.na(.x)  & stringr::str_squish(
               as.character(.x)) != "")) %>%
  mutate(bite_flag = case_when(
    has_child_been_bitten == 1 ~ TRUE,
    has_child_been_bitten == 0 ~ FALSE,
    incident_present ~ TRUE, 
    TRUE ~ NA),
    bite_flag_correct = case_when(
      child_cnt == 0 ~ NA, # households without children
      !is.na(bite_flag) ~ bite_flag,
      TRUE ~ FALSE))
  
# checks

df %>% count(has_child_been_bitten, bite_flag, bite_flag_correct)

df %>% summarise(n_total = n(),
                 n_bite_flag_correct = sum(bite_flag_correct, na.rm = TRUE), 
                 n_incident_present = sum(incident_present, na.rm = TRUE),
                 n_children_bitten = sum(has_child_been_bitten == 1, na.rm = TRUE))
```

## Check duplicate submissions

Create household groups to separate multiple submissions: one row per household.

```{r household-summary}
household_summary <- df %>%  
  group_by(household_id) %>% 
  summarise(n_household_submissions = n(),
            n_bite_rows = sum(bite_flag_correct, na.rm = TRUE),
            .groups = "drop")

summary(household_summary$n_household_submissions)

household_summary %>% count(n_household_submissions, sort = TRUE)

household_summary %>% summarise(
  n_duplicate_households = sum(n_household_submissions > 1))
```

## Create "df_tidy"

Remove duplicate submissions

```{r remove-duplicates}
df_tidy <- df %>% 
  arrange(household_id) %>% 
  group_by(household_id) %>% 
  slice(1) %>% 
  ungroup()

# check 
nrow(df)
nrow(df_tidy)

df_tidy %>% 
  count(household_id, name = "n_household_submissions") %>% 
  summarise(max_household_submissions = max(n_household_submissions))
```

# Variables

## Child count

```{r child-count}
# character to numeric
df_tidy <- df_tidy %>% 
  mutate(child_cnt_raw = as.character(child_cnt), 
         child_cnt_raw = str_to_lower(str_trim(child_cnt_raw)),
         child_cnt = case_when(
           child_cnt_raw %in% c("", "na", "n/a") ~ NA_real_,
           child_cnt_raw %in% c("none", "zero", "0") ~ 0,
           child_cnt_raw %in% c("one", "1") ~ 1,
           child_cnt_raw %in% c("two", "2") ~ 2,
           child_cnt_raw %in% c("three", "3") ~ 3,
           child_cnt_raw %in% c("four", "4") ~ 4,
           child_cnt_raw %in% c("five", "5") ~ 5,
           TRUE ~suppressWarnings(as.numeric(child_cnt_raw))))

# check
df_tidy %>% count(child_cnt, sort = TRUE)

# confirm non-child rows are empty
df_tidy %>% filter(child_cnt == 0) %>% 
  summarise(n_household_submissions = n(),
            n_child_age = sum(!is.na(child_age_at_bite)),
            n_child_id = sum(!is.na(child_id)),
            n_dog_sex = sum(!is.na(dog_sex)))
```

## Non-child rows

eliminate missingness

```{r no-children}
df_no_child <- df_tidy %>% filter(child_cnt == 0)

df_child <- df_tidy %>% filter(child_cnt > 0)

# check
nrow(df_tidy)
nrow(df_no_child)
nrow(df_child)

df_child %>% count(child_cnt, sort = TRUE) # now we use "df_child"
```

## Bite-only dataframe

Create a dataframe for all bite incidents

```{r df-bites}
df_bite <- df_child %>% 
  filter(bite_flag_correct == TRUE)

# check
nrow(df_child)
nrow(df_bite)

df_bite %>% 
  count(bite_flag_correct, sort = TRUE)
```

# Demographic Standardization

Standardize child count, bite flag, child and dog age, child gender, and create a neuter status variable.

## Child age

Convert to months and years.

```{r child-age}
# see all raw ages
df_bite %>% count(child_age_at_bite, sort = TRUE) %>% 
  print(n = 100)

df_bite <- df_bite %>% 
  mutate(
    child_age_raw = str_to_lower(str_squish(as.character(child_age_at_bite))),
    child_age_num = readr::parse_number(child_age_raw),
    child_age_months = case_when(
      is.na(child_age_raw) | child_age_raw == "" ~ NA_real_,
      str_detect(child_age_raw, "month|months|mo\\b") ~ child_age_num,
      str_detect(child_age_raw, "year|years|yr\\b|yo\\b") ~ child_age_num * 12,
      str_detect(child_age_raw, "^around\\s*[0-9]") ~ child_age_num * 12,
      child_age_raw %in% as.character(0:18) ~ child_age_num *12,
      str_detect(child_age_raw, "^[0-9]+\\.[0-9]+$") ~ child_age_num * 12,
                              TRUE ~ NA_real_),
    child_age_years = child_age_months / 12)

# check
df_bite %>%
  count(child_age_at_bite, child_age_months, sort = TRUE)

# drop columns
df_bite <- df_bite %>% select(-child_age_raw, -child_age_num)

# summary
df_bite %>% select(child_age_months, child_age_years) %>% summary()

class(df_bite$child_age_months) 
class(df_bite$child_age_years) # both numeric
```

## Child gender

Standardize character values.

```{r child-gender}
df_bite <- df_bite %>% mutate(child_sex_raw = str_to_lower(str_squish(child_sex)),
                            child_gender_std = case_when(is.na(child_sex_raw) | child_sex_raw == "" ~ NA_character_,
                                                               child_sex_raw %in% c("female", "f", "girl", "woman") ~ "Female",
                                                               child_sex_raw %in% c("male", "boy", "man") ~ "Male",
                                                               str_detect(child_sex_raw, "non[- ]?binary|nb\\b|genderqueer|trans|other") ~ "Other/Nonbinary",
                                                               str_detect(child_sex_raw, "prefer not|prefer not to say|decline") ~ "Prefer not to say",
                                                               TRUE ~ "Needs review"))
# check
df_bite %>% 
  count(child_gender_std, sort = TRUE)

# verify Needs review
df_bite %>% 
  filter(child_gender_std == "Needs review") %>% 
  count(child_sex_raw, sort = TRUE)

# remove columns
df_bite <- df_bite %>%  
  select(-child_sex_raw)

class(df_bite$child_gender_std) # character
```

## Dog age

Convert to months and years.

```{r dog-age}
# inspect raw values
df_bite %>% count(dog_age_at_bite, sort = TRUE) %>% 
  print(n = 100)

df_bite <- df_bite %>% 
  mutate(
    dog_age_raw = str_to_lower(str_squish(as.character(dog_age_at_bite))),
    dog_age_num = readr::parse_number(dog_age_raw),
    dog_age_months = case_when(
      is.na(dog_age_raw) | dog_age_raw == "" ~ NA_real_,
      str_detect(dog_age_raw, "month|months|mo\\b") ~ dog_age_num, # months
      str_detect(dog_age_raw, "year|years|yr\\b|yrs\\b|yo\\b") ~ dog_age_num * 12, # years
      str_detect(dog_age_raw, "^around\\s*[0-9]") ~ dog_age_num * 12, 
      dog_age_raw %in% as.character(0:30)~ dog_age_num *12,
      str_detect(dog_age_raw, "^[0-9]+\\.[0-9]+$") ~ dog_age_num * 12,
      TRUE ~ NA_real_),
    dog_age_years = dog_age_months / 12)

# check
df_bite %>%
  count(dog_age_at_bite, dog_age_months, sort = TRUE)

# drop cols
df_bite <- df_bite %>% 
  select(-dog_age_raw, - dog_age_num)

# summary
df_bite %>% select(dog_age_months, dog_age_years) %>% 
  summary()

class(df_bite$dog_age_months) 
class(df_bite$dog_age_years) # numeric
```

## Dog sex/neuter status

Inspect dog gender to formalize dog_neuter status variable.

```{r neuter-status}
# inspect raw values 
df_bite %>% 
  count(dog_sex, sort = TRUE)

# create variable for sterilization
df_bite <- df_bite %>% 
  mutate(
    dog_sex_raw = str_to_lower(str_squish(dog_sex)),
    dog_sex_std = case_when(
      is.na(dog_sex_raw) | dog_sex_raw == "" ~ NA_character_,
      str_detect(dog_sex_raw, "\\bfemale\\b") ~ "Female",
      str_detect(dog_sex_raw, "\\bmale\\b") ~ "Male",
      TRUE ~ "Needs review"),
    dog_neuter = case_when(
      is.na(dog_sex_raw) | dog_sex_raw == "" ~ NA,
      str_detect(dog_sex_raw, "spayed|neutered") ~ TRUE,
      dog_sex_raw %in% 
        c("male", "female") ~ FALSE, TRUE ~ NA))
                                           
# check
df_bite %>%  count(dog_sex, dog_sex_std, dog_neuter, sort = TRUE)

df_bite %>% 
  filter(dog_sex_std == "Needs review") %>% 
  count(dog_sex_raw, sort = TRUE)

class(df_bite$dog_neuter) # logical

# drop cols
df_bite <- df_bite %>% 
  select(-dog_sex_raw)
```

# Column standardization

Identify and save survey items with multiple checkbox items.

```{r checkbox-cols}

# All checkbox items
checkbox_cols <- names(df_bite)[str_detect(names(df_bite), "__")]

# "Other" write-in items
checkbox_other <- checkbox_cols[str_detect(
  checkbox_cols,
  "__other($|_)|other_text|specify|describe")]

# save and convert selected checkbox columns to logical
checkbox_logical <- setdiff(checkbox_cols, checkbox_other)

# check
checkbox_cols
checkbox_other
checkbox_logical
```

## Checkbox variables

convert checkbox items to logical (T/F).

```{r checkbox-conversion}
# unchecked = FALSE; checked = TRUE
df_bite <- df_bite %>%
  mutate(across(all_of(checkbox_logical), ~{
    x <- .x
    if (is.logical(x)) { 
      replace(x, is.na (x), FALSE)
      } else if (is.numeric(x) || is.integer(x)) {
        case_when(
          x == 1 ~ TRUE,
          x == 0 ~ FALSE,
          is.na(x) ~ FALSE,
          TRUE ~ FALSE)
        } else if (is.character(x)) {
          x2 <- str_squish(as.character(x))
          case_when(
            is.na(x2) | x2 == "" ~ FALSE,
            TRUE ~ TRUE )
          } else { 
            FALSE }})) # all others

# retain write-in "other" response as character string
df_bite <- df_bite %>% mutate(across(all_of(checkbox_other), ~{
    x <- str_squish(as.character(.x))
    na_if(x, "")}))

# check
table(sapply(df_bite[checkbox_logical], class))
table(sapply(df_bite[checkbox_other], class))

df_bite %>% count(triggered_bite__hugged_the_dog, sort = TRUE)

# check logical missingness
df_bite %>% summarise(across(starts_with(
  "dog_behavior_prior_to_bite__"),
  ~ paste(unique(as.character(.x)), 
          collapse = " | "))) %>%
  print(width = Inf)

df_bite %>% summarise(across(starts_with(
  "dog_behavior_prior_to_bite__"),
  ~ sum(.x == TRUE, na.rm = TRUE))) %>%
  pivot_longer(everything(),
               names_to = "warning_item",
               values_to = "n_true") %>%
  arrange(desc(n_true))

dog_warning_cols <- names(df_bite) %>%
  stringr::str_subset("^dog_behavior_prior_to_bite__")

# convert true to logical
df_bite <- df_bite %>% mutate(across(all_of(setdiff(
  dog_warning_cols, 
  "dog_behavior_prior_to_bite__other")),
  ~ tidyr::replace_na(.x, FALSE)),
  dog_warning__other_clean = stringr::str_to_lower(
    stringr::str_squish(
      dog_behavior_prior_to_bite__other)),
  dog_behavior_prior_to_bite__none = if_else(str_detect(
    dog_warning__other_clean, 
    "didn.?t see|did not see|alone|no warning|no forewarning|without warning"), 
    TRUE,
    dog_behavior_prior_to_bite__none))

# re-build columns 
dog_warning_cols <- dog_warning_cols %>% setdiff(c(
  "dog_behavior_prior_to_bite__none",
  "dog_behavior_prior_to_bite__other"))

df_bite <- df_bite %>% mutate(dog_warning_any = if_any(all_of(
  dog_warning_cols),
    ~ .x == TRUE),
  dog_warning_none = dog_behavior_prior_to_bite__none == TRUE,
  dog_warning_count = rowSums(across(all_of(
    dog_warning_cols), 
    ~ as.numeric(.x)),
    na.rm = TRUE))

# check
names(df_bite)[startsWith(names(df_bite), 
                          "dog_behavior_prior_to_bite__")] %in% checkbox_logical
```

## Check variables

List all variables.

```{r identify-variables}

df_bite %>% select(where(~ !is.logical(.))) %>% names()
```

# Build columns

Build logical columns of behavioral variables.

## dog warning

Build dog_warning_any, \_none, \_count variables.

```{r dog-warning}
# categorize "others" 
df_bite <- df_bite %>% mutate(
  dog_warning__other_clean = stringr::str_to_lower(
    stringr::str_squish(dog_behavior_prior_to_bite__other)),
  dog_behavior_prior_to_bite__air_snapped = if_else(
    stringr::str_detect(dog_warning__other_clean, 
                        "air snap|snapped|snap"),
    TRUE, 
    dog_behavior_prior_to_bite__air_snapped,
    missing = dog_behavior_prior_to_bite__air_snapped),
  dog_behavior_prior_to_bite__bared_teeth_lifted_lips = if_else(
    stringr::str_detect(dog_warning__other_clean, 
                        "bared teeth|showed teeth|lifted lips|lip"),
    TRUE,
    dog_behavior_prior_to_bite__bared_teeth_lifted_lips,
    missing = dog_behavior_prior_to_bite__bared_teeth_lifted_lips),
  dog_behavior_prior_to_bite__barked = if_else(
    stringr::str_detect(dog_warning__other_clean,
                        "bark|barked|barking"),
    TRUE,
    dog_behavior_prior_to_bite__barked,
    missing = dog_behavior_prior_to_bite__barked),
  dog_behavior_prior_to_bite__cowered_away = if_else(
    stringr::str_detect(dog_warning__other_clean, 
                        "cower|cowered|backed away|moved away|retreated|hid"),
    TRUE,
    dog_behavior_prior_to_bite__cowered_away,
    missing = dog_behavior_prior_to_bite__cowered_away),
  dog_behavior_prior_to_bite__growled = if_else(
    stringr::str_detect(dog_warning__other_clean, 
                        "growl|growled|growling"),
    TRUE,
    dog_behavior_prior_to_bite__growled,
    missing = dog_behavior_prior_to_bite__growled), 
  dog_behavior_prior_to_bite__lunged = if_else(
    stringr::str_detect(dog_warning__other_clean, 
                        "lunge|lunged|charged"),
    TRUE,
    dog_behavior_prior_to_bite__lunged,
    missing = dog_behavior_prior_to_bite__lunged),
  dog_behavior_prior_to_bite__raised_hackles = if_else(
    stringr::str_detect(dog_warning__other_clean, 
                        "hackles|hair raised|raised hair"),
    TRUE,
    dog_behavior_prior_to_bite__raised_hackles,
    missing = dog_behavior_prior_to_bite__raised_hackles),
  dog_behavior_prior_to_bite__tucked_tail = if_else(
    stringr::str_detect(dog_warning__other_clean, 
                        "tucked tail|tail tucked"),
    TRUE,
    dog_behavior_prior_to_bite__tucked_tail,
    missing = dog_behavior_prior_to_bite__tucked_tail),
  dog_behavior_prior_to_bite__whale_eye = if_else(
    stringr::str_detect(dog_warning__other_clean, 
                        "whale eye|side eye|showing whites|white of"),
    TRUE,
    dog_behavior_prior_to_bite__whale_eye,
    missing = dog_behavior_prior_to_bite__whale_eye),
  dog_behavior_prior_to_bite__none = if_else(
    stringr::str_detect(dog_warning__other_clean, 
                        "no warning|no forewarning|without warning|none"),
    TRUE,
    dog_behavior_prior_to_bite__none,
    missing = dog_behavior_prior_to_bite__none))

# re-build columns
dog_warning_cols <- dog_warning_cols %>% 
  setdiff(c("dog_behavior_prior_to_bite__none", "dog_behavior_prior_to_bite__other"))

# inspect
dog_warning_cols

# variable summary
df_bite <- df_bite %>% mutate(
  dog_warning_any = if_any(all_of(dog_warning_cols), ~ .x == TRUE),
  dog_warning_none = dog_behavior_prior_to_bite__none == TRUE,
  dog_warning_count = rowSums(across(all_of(
    dog_warning_cols), 
    ~ as.numeric(.x)),
    na.rm = TRUE))

# check
df_bite %>% count(dog_warning_any)

df_bite %>% count(dog_warning_count)

df_bite %>% summarise(across(all_of(
  dog_warning_cols), 
  ~ sum(.x, 
        na.rm = TRUE))) %>%
  pivot_longer(
    everything(),
    names_to = "dog_warning",
    values_to = "count") %>% 
  arrange(desc(count))
```

## dog triggers

Build dog_trigger_any, \_none, \_count variables.

```{r dog-triggers}
# categorize others
df_bite <- df_bite %>% mutate(
  triggered_bite__other_clean = 
    stringr::str_to_lower(stringr::str_squish(triggered_bite__other)),
  triggered_bite__pet_legs_paws = if_else(
    str_detect(triggered_bite__other_clean, 
               "leg|foot|paw"),
    TRUE,
    triggered_bite__pet_legs_paws),
  triggered_bite__approached_dog_while_eating = if_else(
    str_detect(triggered_bite__other_clean, 
               "food|pizza|bread|treat|eating"),
    TRUE,
    triggered_bite__approached_dog_while_eating),
  triggered_bite__approached_dog_while_sleeping = if_else(
    str_detect(triggered_bite__other_clean,
               "sleep|sleeping|middle of night"),
    TRUE,
    triggered_bite__approached_dog_while_sleeping),
  triggered_bite__stepped_fell_on_dog = if_else(
    str_detect(triggered_bite__other_clean, 
               "stomped|stepped|fell|fall|knocking over"),
    TRUE,
    triggered_bite__stepped_fell_on_dog),
  triggered_bite__child_running = if_else(
    str_detect(triggered_bite__other_clean, 
               "running|ran"),
    TRUE,
    triggered_bite__child_running),
  triggered_bite__ran_at_dog = if_else(
    str_detect(triggered_bite__other_clean, 
               "ran toward dog|ran at dog"),
    TRUE,
    triggered_bite__ran_at_dog),
  triggered_bite__pet_head = if_else(
    str_detect(triggered_bite__other_clean, "pet dog on head|head"),
    TRUE,
    triggered_bite__pet_head),
  triggered_bite__no_obvious_trigger = if_else(
    str_detect(triggered_bite__other_clean, "didn.?t see|did not see|no obvious"),
    TRUE,
    triggered_bite__no_obvious_trigger),
  # New categories
  triggered_bite__redirected_aggression = str_detect(
    triggered_bite__other_clean,
    "redirected|another dog"),
  triggered_bite__resource_conflict = str_detect(
    triggered_bite__other_clean,
    "steal|take|treat|food|pizza|bread"),
  triggered_bite__tight_space_trapped = str_detect(
    triggered_bite__other_clean,
    "tight spot|trapped|small space"),
  triggered_bite__moved_dog = str_detect(
    triggered_bite__other_clean,
    "move dog|move the dog"),
  triggered_bite__pet_back = str_detect(
    triggered_bite__other_clean,
    "pet dog.?s back|pet.*back"),
  triggered_bite__face_close = str_detect(
    triggered_bite__other_clean,
    "face.*close|close to dog.?s face"),
  triggered_bite__tail_pull = str_detect(
    triggered_bite__other_clean,
    "tail"),
  triggered_bite__teasing = str_detect(
    triggered_bite__other_clean,
    "teasing|tease"),
  triggered_bite__loud_or_high_pitch = str_detect(
    triggered_bite__other_clean,
    "high pitch|crying|screamed|baby crying"))

# build trigger col
dog_trigger_cols <- names(df_bite) %>%
  stringr::str_subset("^triggered_bite__") %>% setdiff(c(
    "triggered_bite__other",
    "triggered_bite__other_clean"))

df_bite <- df_bite %>% mutate(
  dog_trigger_any = if_any(all_of(
    dog_trigger_cols),
    ~ .x == TRUE),
  dog_trigger_count = rowSums(across(all_of(
    dog_trigger_cols),
    ~ as.numeric(.x)),
    na.rm = TRUE))

# check
df_bite %>% count(dog_trigger_any)

df_bite %>% count(dog_trigger_count)

df_bite %>% summarise(across(all_of(
  dog_trigger_cols),
  ~ sum(.x, na.rm = TRUE))) %>% pivot_longer(everything(),
                                             names_to = "dog trigger",
                                             values_to = "count") %>%
  arrange(desc(count))
```

## bite severity

Build numeric bite_severity_num and bite_severity_clean to include recoded "other" values.

```{r bite-severity}
# inspect
df_bite %>% count(bite_severity, sort = TRUE) %>%
  print(n = Inf, width = Inf)

df_bite <- df_bite %>% mutate(
  bite_severity_num = stringr::str_extract(
    bite_severity,
    "(?<=Score )\\d") %>%
    as.numeric())

# checks
df_bite %>%
  count(bite_severity_num, sort = TRUE)

df_bite %>%
  select(bite_severity, bite_severity_num, bite_severity__other) %>%
  filter(is.na(bite_severity_num) | !is.na(bite_severity__other)) %>%
  print(n = Inf, width = Inf)

# categorize "other"
df_bite <- df_bite %>% mutate(
  bite_severity_other_recoded = case_when(
    str_detect(
      bite_severity__other,
      "puncture|laceration") ~ 3,
    str_detect(
      bite_severity__other,
      "no mark on the skin|no visible injury|clothes") ~ 1,
    TRUE ~ NA_real_),
  bite_severity_clean = coalesce(
    bite_severity_num,
    bite_severity_other_recoded))

# check 
df_bite %>% count(bite_severity_clean, sort = TRUE)

df_bite %>% select(
  bite_severity,
  bite_severity__other,
  bite_severity_num,
  bite_severity_other_recoded,
  bite_severity_clean) %>%
  filter(!is.na(bite_severity__other) | is.na(bite_severity_clean)) %>%
  print(n = Inf, width = Inf)

# remove helper
df_bite <- df_bite %>% mutate(bite_severity = bite_severity_clean) %>%
  select(-bite_severity_num, -bite_severity_other_recoded, -bite_severity_clean)
```

## dog personality

Build dog_personality_any (logical) and dog_personality_count (numeric).

```{r dog-personality}
# inspect "other"
df_bite %>% filter(!is.na(dog_personality__other)) %>% select(
  dog_personality__other) %>% distinct() %>%
  print(n = Inf, width = Inf)

# existing categories
dog_personality_cols

# include others in new categories
df_bite <- df_bite %>% mutate(
  dog_personality__other_clean = 
    str_to_lower(str_squish(dog_personality__other)),
  dog_personality__newly_adopted = str_detect(
    dog_personality__other_clean,
    "first day|hadn.?t had|rescued|bad situation"),
  dog_personality__work_focused = str_detect(
    dog_personality__other_clean,
    "work minded|working"),
  dog_personality__cognitive_decline = str_detect(
    dog_personality__other_clean,
    "dementia"),
  dog_personality__dog_selective = str_detect(
    dog_personality__other_clean,
    "dog selective"))


dog_personality_cols <- names(df_bite) %>%
  stringr::str_subset("^dog_personality__") %>%
  setdiff(c("dog_personality__other", "dog_personality__other_clean"))

# check
dog_personality_cols

# build summary variables
df_bite <- df_bite %>% mutate(
  dog_personality_any = if_any(all_of(
    dog_personality_cols),
    ~ .x == TRUE),
  dog_personality_count = rowSums(
    across(
      all_of(dog_personality_cols),
      ~ as.numeric(.x)),
    na.rm = TRUE))

# checks
df_bite %>% count(dog_personality_any)

df_bite %>% count(dog_personality_count)

df_bite %>% summarise(across(all_of(
  dog_personality_cols),
  ~ sum(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(),
               names_to = "personality_trait",
               values_to = "count") %>%
  arrange(desc(count))
```

## child personality

Build child_personality_any (logical) and child_personality_count (numeric).

```{r child-personality}
# inspect
child_personality_cols

df_bite %>% 
  filter(!is.na(personality_at_bite__other)) %>% 
  select(personality_at_bite__other) %>% 
  distinct() %>% 
  print(n = Inf, width = Inf)

# categorize "other" 
df_bite <- df_bite %>% mutate(
  personality_at_bite__other_clean = stringr::str_to_lower(
    stringr::str_squish(personality_at_bite__other)),
  personality_at_bite__grabby = stringr::str_detect(
    personality_at_bite__other_clean,
    "grabby|grab|grabbing"),
  personality_at_bite__disabled = stringr::str_detect(
    personality_at_bite__other_clean,
    "handicapped|disabled|disability"),
  personality_at_bite__disobedient = stringr::str_detect(
    personality_at_bite__other_clean,
    "not very obedient|disobedient|not obedient")) # disobedient might be a mistake (not dog)

# identify 
child_personality_cols <- names(df_bite) %>%
  stringr::str_subset("^personality_at_bite__") %>%
  setdiff(c(
    "personality_at_bite__other",
    "personality_at_bite__other_clean"))

child_personality_cols

# build summary variables
df_bite <- df_bite %>% mutate(
  child_personality_any = if_any(all_of(
    child_personality_cols), 
    ~ .x == TRUE),
  child_personality_count = rowSums(across(all_of(
    child_personality_cols),
    ~ as.numeric(.x)),
    na.rm = TRUE))

# checks
df_bite %>% count(child_personality_any)

df_bite %>% count(child_personality_count)

df_bite %>% summarise(across(all_of(
  child_personality_cols),
  ~ sum(.x, na.rm = TRUE))) %>%
  pivot_longer(
    everything(),
    names_to = "child_personality_trait",
    values_to = "count") %>% arrange(desc(count))
```

## approach direction \<- child action prior to bite

Create approach_direction (categorical) column creating child_approached_dog, dog_approached_child, mutual_approach, and unknown.

```{r approach-direction}
# inspect values
df_bite %>% count(desc_action_prior_to_bite, sort = TRUE) %>% 
  print(n = 32, width = Inf)

# std interaction direction
df_bite <- df_bite %>% mutate(
  approach_direction = case_when(
    desc_action_prior_to_bite == "The child approached the dog" ~
      "child_approached_dog",
    desc_action_prior_to_bite == "The dog approached the child" ~
      "dog_approached_child",
    desc_action_prior_to_bite == "They approached each other" ~
      "mutual_approach",
    desc_action_prior_to_bite == "I'm not sure" ~
      "unknown",
    TRUE ~ NA_character_))

# check
df_bite %>% count(approach_direction, sort = TRUE)

df_bite %>% select(desc_action_prior_to_bite) %>% 
  distinct() %>% 
  print(n = 32, width = Inf)
```

# Data checks

## Dimensions

Assess consistency.

```{r dimensions-check}
dim(df_bite)

stopifnot(nrow(df_bite) == 32)
```

## Structure

Check ages and counts are numeric, checkboxes are logical, check categories.

```{r structure-check}
glimpse(df_bite)
```

## Missings

Check NAs.

```{r missings-check}
missing_summary <- df_bite %>% 
  summarise(across(everything(), 
                   ~ sum(is.na(.x)))) %>%
  pivot_longer(everything(), 
               names_to = "variable",
               values_to = "n_missing") %>% 
  arrange(desc(n_missing))

missing_summary
```

## Variables

Check numeric, logical, and categorical variables.

```{r variables-check}

# numeric
df_bite %>% select(where(is.numeric)) %>%
  summary()

# logical checkbox vars
df_bite %>% select(where(is.logical)) %>%
  summarise(across(everything(),
                   ~ sum(.x, na.rm = TRUE))) %>%
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "TRUE_count") %>%
  arrange(desc(TRUE_count))

# character 
names(df_bite)[sapply(df_bite, is.character)]

# all 
df_bite %>% select(ends_with("_any")) %>% 
  summary()

df_bite %>% select(ends_with("_count")) %>% 
  summary()

# bite severity
table(df_bite$bite_severity, useNA = "ifany")

# child age
summary(df_bite$child_age_years)

summary(df_bite$child_age_months)

# dog age
summary(df_bite$dog_age_years)

summary(df_bite$dog_age_months)

# child gender
table(df_bite$child_gender_std, useNA = "ifany")

# dog sex
table(df_bite$dog_sex_std, useNA = "ifany")

# reproductive status
table(df_bite$dog_neuter, useNA = "ifany")

# approach direction
table(df_bite$approach_direction, useNA = "ifany")
```

## Remove helper vars

```{r remove-helpers}
df_bite <- df_bite %>% select(
  -child_cnt_raw,
  -dog_warning__other_clean,
  -triggered_bite__other_clean,
  -dog_personality__other_clean,
  -personality_at_bite__other_clean)

# check
names(df_bite) %>% stringr::str_subset(
  "_clean$|_other_clean$|_num$|_raw$")
```

## Duplicates

```{r duplicates-check}
df_bite %>% count(household_id) %>% 
  filter(n > 1)
```

# Save as RDS

Save final dataframe

```{r save-RDS}
# save as df_clean.rds
saveRDS(df_bite, "../data/df_clean.rds")
```
