---
title: "Exploratory Data Analysis"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo=TRUE)
```

# Overview

This notebook documents the exploratory data analysis undertaken for this project.

# Loading the Data

Load the tidy data set into memory. If it does not exist, then run the notebook that builds it.

```{r}
tidy_data_filepath <- '../data/tidy.Rds'
stopifnot(file.exists(tidy_data_filepath))

df <- readRDS(tidy_data_filepath)
stopifnot(nrow(df) == 192) # accurate number or responders 
print(df, n=6)
```

# Summarize Data

```{r}
summary(df)
```

```{r}
# Number of unique owners.
length(unique(df$owner_id))
```

# Display Session Info

```{r}
utils::sessionInfo()
```

# Load the Standardized Data

```{r}
df_bite <- read_csv("../data/bite-data-std.csv")

# cleaning check
dim(df_bite)
```

# Child Demographics

## Child Gender

Variable "child_gender_std" is a categorical variable indicating the reported gender of the child involved in bite incident.

```{r child-gender}
# confirm std categories 
unique(df_bite$child_gender_std)

df_bite %>% count(child_gender_std, sort = TRUE)

# check missing
df_bite %>% summarise(sum(is.na(child_gender_std)))

# summary tables (gender frequency)
df_bite %>% count(child_gender_std) %>% 
  mutate(percent = round(n / sum(n) * 100, 1))

# visual 
## bar chart
ggplot(df_bite, aes(x = child_gender_std)) + 
  geom_bar() + 
  labs(title = "Child Gender Distributiuon (Bite Incidents)",
       x = "Child Gender",
       y = "Count") + 
  theme_minimal()
```

## Child Age

Variable "child_age_months" and "child_age_years" are continuous variables indicating the age of child involved, expressed in months and years.

```{r child-age}
# inspect values
summary(df_bite$child_age_months)
summary(df_bite$child_age_years)

# check missing
df_bite %>% summarise(
  n_child_months_missing = sum(is.na(child_age_months)),
  n_child_years_missing = sum(is.na(child_age_years)))

# summary tables 
## MONTHS
df_bite %>% summarise(
  n_child_months = sum(!is.na(child_age_months)),
  mean_child_months = mean(child_age_months, na.rm = TRUE),
  sd_child_months = sd(child_age_months, na.rm = TRUE),
  min_child_months = min(child_age_months, na.rm = TRUE),
  max_child_months = max(child_age_months, na.rm = TRUE),
  median_child_months = median(child_age_months, na.rm = TRUE))

## YEARS
df_bite %>% summarise(
  n_child_years = sum(!is.na(child_age_years)),
  mean_child_years = mean(child_age_years, na.rm = TRUE),
  sd_child_years = sd(child_age_years, na.rm = TRUE),
  min_child_years = min(child_age_years, na.rm = TRUE),
  max_child_years = max(child_age_years, na.rm = TRUE),
  median_child_years = median(child_age_years, na.rm = TRUE))

# visuals 
## histogram
ggplot(df_bite, aes(x = child_age_years)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Child Age Distribution (Bite Incidents)",
       x = "Child Age (Years)",
       y = "Count") +
  theme_minimal()

## boxplot
ggplot(df_bite, aes(y = child_age_years)) +
  geom_boxplot() +
  labs(title = "Child Age Boxplot (Bite Incidents)",
       y = "Child Age (Years)") +
  theme_minimal()
```

## Child Count (per household)

Variable "child_cnt_raw" and "child_cnt" are variables indicating number of children reported in the household.

```{r child-count}
# verify equivalence
sum(df_bite$child_cnt != df_bite$child_cnt_raw, na.rm = TRUE)

# check structure
str(df_bite$child_cnt_raw)
str(df_bite$child_cnt) # both numerical, ignore child_cnt_raw moving forward

# inspect values
summary(df_bite$child_cnt)

df_bite %>% summarise(
  n_child_cnt_missing = sum(is.na(child_cnt)))

# summary statistics
df_bite %>% summarise(
  n_child_cnt = sum(!is.na(child_cnt)),
  mean_child_cnt = mean(child_cnt, na.rm = TRUE),
  sd_child_cnt = sd(child_cnt, na.rm = TRUE), 
  min_child_cnt = min(child_cnt, na.rm = TRUE),
  max_child_cnt = max(child_cnt, na.rm = TRUE),
  median_child_cnt = median(child_cnt, na.rm = TRUE))

# summary table
df_bite %>% count(child_cnt) %>% mutate(
  percent = round(n / sum(n) * 100, 1))

# Visuals
## barplot within incidents 
ggplot(df_bite, aes(x = factor(child_cnt))) +
  geom_bar() +
  labs(title = "Children Per Household (Bite Incident)",
       x = "Children",
       y = "n") +
  theme_minimal()

## barplot total sample size to guage distribution ***BROKEN***
ggplot(df_bite, aes(x = factor(child_cnt))) +
  geom_bar() +
  labs(title = "Children Per Household",
       x = "Children",
       y = "n") +
  theme_minimal()
```

# Dog Demographics

## Dog sex

Variable "dog_sex_std" is a categorical variable indicating the reported sex of the dog involved in bite incident, independent of neuter status.

```{r dog-sex}
# confirm categories
unique(df_bite$dog_sex_std)

df_bite %>% count(dog_sex_std, sort = TRUE)

# check missing
df_bite %>% summarise(sum(is.na(dog_sex_std)))

# summary tables
df_bite %>% count(dog_sex_std) %>% mutate(
  percent = round(n / sum(n) * 100, 1))

# Visuals
## barplot
ggplot(df_bite, aes(x = dog_sex_std)) +
  geom_bar() +
  labs(title = "Dog Sex Distribution (Bite Incidents)",
       x = "Sex",
       y = "n") +
  theme_minimal()
```

## Neuter status

Variable "dog_neuter" is a logical variable indicating whether the dog involved in the bite incident was sterilized (TRUE = neutered/spayed, FALSE = intact).

```{r sterilization}
# inspect values
unique(df_bite$dog_neuter)

df_bite %>% count(dog_neuter, sort = TRUE)

# check missing 
df_bite %>% summarise(sum(is.na(dog_neuter)))

# summary table
df_bite %>% count(dog_neuter) %>% mutate(
  percent = round(n / sum(n) * 100, 1))

# visual
## barplot
ggplot(df_bite, aes(x = factor(dog_neuter, 
                               levels = c(FALSE, TRUE),
                               labels = c("Intact", "Sterilized")))) +
  geom_bar() + 
  labs(title = "Dog Sterilization Status (Bite Incidents)",
       x = "Sterilized",
       y = "n") +
  theme_minimal()
```

# make double column

## gender and neuter-status

## Dog Age

Variable "dog_age_months" and "dog_age_years" are continuous variables indicating the age of the dog involved in the bite incident.

```{r dog-age}
# inspect values
summary(df_bite$dog_age_months)
summary(df_bite$dog_age_years)

# check missing
df_bite %>% summarise(sum(is.na(dog_age_months))) 
df_bite %>% summarise(sum(is.na(dog_age_years))) 

# summary tables
## MONTHS
df_bite %>% summarise(
  n_dog_months = sum(!is.na(dog_age_months)),
  mean_dog_months = mean(dog_age_months, na.rm = TRUE),
  sd_dog_months = sd(dog_age_months, na.rm = TRUE),
  min_dog_months = min(dog_age_months, na.rm = TRUE),
  max_dog_months = max(dog_age_months, na.rm = TRUE),
  median_dog_months = median(dog_age_months, na.rm = TRUE))

## YEARS  
df_bite %>% summarise(
  n_dog_years = sum(!is.na(dog_age_years)),
  mean_dog_years = mean(dog_age_years, na.rm = TRUE),
  sd_dog_years = sd(dog_age_years, na.rm = TRUE),
  min_dog_years = min(dog_age_years, na.rm = TRUE),
  max_dog_years = max(dog_age_years, na.rm = TRUE),
  median_dog_years = median(dog_age_years, na.rm = TRUE))

# Visuals
## histogram
ggplot(df_bite, aes(x = dog_age_years)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Dog Age Distribution (Bite Incidents)",
       x = "Dog Age (Years)",
       y = "n") +
  theme_minimal()

## boxplot
ggplot(df_bite, aes(y = dog_age_years)) +
  geom_boxplot() +
  labs(title = "Dog Age Boxplot (Bite Incidents)",
       y = "Dog Age (Years)") +
  theme_minimal() 
```

# Demographics Summary Table

## Values table

```{r}
bite_demo_values_table <- tibble(Variable = c(
  "Child gender: Male",
  "Child gender: Female", 
  "Child gender: Prefer not to say",
  
  "Child age (years)",
  
  "Child count per household", 
  
  "Dog sex: Male",
  "Dog sex: Female",
  
  "Dog sterilized",
  "Dog intact",
  
  "Dog age (years)"),
  
  Value = c(
    sum(df_bite$child_gender_std == "Male", na.rm = TRUE),
    sum(df_bite$child_gender_std == "Female", na.rm = TRUE),
    sum(df_bite$child_gender_std == "Prefer not to say", na.rm = TRUE),
                
    mean(df_bite$child_age_years, na.rm = TRUE),
                
    mean(df_bite$child_cnt, na.rm = TRUE),
                
    sum(df_bite$dog_sex_std == "Male", na.rm = TRUE),
    sum(df_bite$dog_sex_std == "Female", na.rm = TRUE),
                
                sum(df_bite$dog_neuter == TRUE, na.rm = TRUE),
                sum(df_bite$dog_neuter == FALSE, na.rm = TRUE),
                
                mean(df_bite$dog_age_years, na.rm = TRUE)))

bite_demo_values_table
```

## Demographic table with counts and proportions

```{r}
N <- nrow(df_bite)

bite_demo_table <- tibble(Variable = c(
  "Child gender: Male",
  "Child gender: Female", 
  "Child gender: Prefer not to say",
  
  "Child age (years)",
  
  "Child count per household", 
  
  "Dog sex: Male",
  "Dog sex: Female",
  
  "Dog sterilized",
  "Dog intact",
  
  "Dog age (years)"),
  
  Value = c(
    sprintf("%d (%.1f%%)",
            sum(df_bite$child_gender_std == "Male", na.rm = TRUE),
            100 * sum(df_bite$child_gender_std == "Male", na.rm = TRUE) / N), 
    sprintf("%d (%.1f%%)",
            sum(df_bite$child_gender_std == "Female", na.rm = TRUE),
            100 * sum(df_bite$child_gender_std == "Female", na.rm = TRUE) / N),
    sprintf("%d (%.1f%%)",
            sum(df_bite$child_gender_std == "Prefer not to say", na.rm = TRUE),
            100 * sum(df_bite$child_gender_std == "Prefer not to say", na.rm = TRUE) / N), 
    
    sprintf("%.2f (%.2f)",
            mean(df_bite$child_age_years, na.rm = TRUE),
            sd(df_bite$child_age_years, na.rm = TRUE)),
    
    sprintf("%.2f (%.2f)",
            mean(df_bite$child_cnt, na.rm = TRUE),
            sd(df_bite$child_cnt, na.rm = TRUE)),
    
    sprintf("%d (%.1f%%)",
            sum(df_bite$dog_sex_std == "Male", na.rm = TRUE),
            100 * sum(df_bite$dog_sex_std == "Male", na.rm = TRUE) / N), 
    sprintf("%d (%.1f%%)",
            sum(df_bite$dog_sex_std == "Female", na.rm = TRUE),
            100 * sum(df_bite$dog_sex_std == "Female", na.rm = TRUE) / N),
    
    sprintf("%d (%.1f%%)",
            sum(df_bite$dog_neuter == TRUE, na.rm = TRUE),
            100 * sum(df_bite$dog_neuter == TRUE, na.rm = TRUE) / N), 
    sprintf("%d (%.1f%%)",
            sum(df_bite$dog_neuter == FALSE, na.rm = TRUE),
            100 * sum(df_bite$dog_neuter == FALSE, na.rm = TRUE) / N), 
    
    sprintf("%.2f (%.2f)",
            mean(df_bite$dog_age_years, na.rm = TRUE),
            sd(df_bite$dog_age_years, na.rm = TRUE))))

  
bite_demo_table
        
```

# Behavioral Predictors
## prior behavior and severity
```{r approach-direction}
df_bite <- df_bite %>% mutate(
  child_action_category = as.character(child_action_category),
  child_action_category = case_when(
    child_action_category == "child_approached_dog" ~ "child_approached_dog",
    child_action_category == "dog_approached_child" ~ "dog_approached_child",
    child_action_category == "mutual_approach" ~ "mutual_approach",
    child_action_category == "unknown" ~ "unknown",
    TRUE ~ NA_character_),
  child_action_category = factor(
    child_action_category, levels = c(
      "child_approached_dog",
      "dog_approached_child",
      "mutual_approach",
      "unknown")))

# visual
ggplot(df_bite, aes(
  x = child_action_category,
  fill = factor(bite_severity_clean))) + 
  geom_bar(position = "stack", na.rm = TRUE) +
  scale_x_discrete(
    drop = FALSE,
    labels = c(
      child_approached_dog = "Child approached dog",
```


```{r}
dog_approached_child = "Dog approached child",
      mutual_approach = "Mutual approach",
      unknown = "Unknown")) +
  labs(
    title = "Bite Severity by Approach Direction",
    x = "Direction",
    y = "Count",
    fill = "Severity") +
  theme_minimal()
```

## warning signs and bite severity
```{r warning-signals}
df_bite %>% count(dog_warning_any, bite_severity_clean)

# visual
ggplot(df_bite, aes(
  x = factor(bite_severity_clean),
  fill = dog_warning_any)) + 
  geom_bar(position = "stack") + 
  scale_fill_discrete(labels = c(
    "FALSE" = "No warning signal",
    "TRUE" = "Warning signal")) +
  labs(
    title = "Warning Signals by Bite Severity",
    x = "Bite Severity",
    y = "Number of Warning Signals") +
  theme_minimal()

# warning count trend by severity
df_bite %>% group_by(bite_severity_clean) %>% 
  summarise(
    warning_count_mean = mean(dog_warning_count, na.rm = TRUE),
    warning_count_median = median(dog_warning_count, na.rm = TRUE),
    n = n())

# warning count distribution 
ggplot(df_bite, aes(
  x = factor(bite_severity_clean),
  fill = factor(dog_warning_count))) + 
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) + 
  labs(
    title = "Warning Signal Distribution Across Severity",
    x = "Bite Severity",
    y = "Proportion of Cases",
    fill = "Warning Signal Count") + 
  theme_minimal()

# Jitter plot
ggplot(df_bite,
       aes(
         x = factor(bite_severity_clean),
         y = dog_warning_count)) +
  geom_jitter(
    width = 0.15,
    height = 0.05,
    alpha = 0.7,
    size = 3) +
  labs(
    title = "Warning Signal Across Bite Severities",
    x = "Bite Severity",
    y = "Warning Signal Count") +
  theme_minimal()
```

## approach direction x warning signals
```{r}
df_bite %>% count(child_action_category, dog_warning_any)

# visual 
ggplot(df_bite, aes(
  x = child_action_category,
  fill = dog_warning_any)) +
  geom_bar(position = "fill") + 
  scale_y_continuous(labels = scales::percent) +
  scale_fill_discrete(labels = c(
    "FALSE" = "No warning signal",
    "TRUE" = "Warning Signal")) +
  scale_x_discrete(labels = c(
    child_approached_dog = "Child approached dog",
    dog_approached_child = "Dog approached child",
    mutual_approach = "Mutual approach",
    unknown = "Unknown")) +
  labs(
    title = "Warning Signals by Approach Direction",
    x = "Approach Direction",
    y = "Proportion of Bite Cases",
    fill = "Warning Signal") + 
  theme_minimal()

# proportionality visual

# define warning rate
warning_summary <- df_bite %>%
  group_by(child_action_category) %>%
  summarise(
    warning_rate = mean(dog_warning_any, na.rm = TRUE),
    n = n())

# visual
ggplot(warning_summary, aes(
  x = child_action_category,
  y = warning_rate)) +
  geom_point(size = 4) +
  geom_segment(aes(
    x = child_action_category,
    xend = child_action_category,
    y = 0,
    yend = warning_rate)) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)) +
  scale_x_discrete(labels = c(
    child_approached_dog = "Child approached dog",
    dog_approached_child = "Dog approached child",
    mutual_approach = "Mutual approach",
    unknown = "Unknown")) +
  labs(
    title = "Proportion of Bite Incidents With Warning Signals",
    x = "Interaction Direction",
    y = "Warning Signal Prevalence") +
  theme_minimal()
```

## child age x severity
```{r}
df_bite %>% group_by(bite_severity_clean) %>% 
  summarise(
    child_age_mean = mean(child_age_years, na.rm = TRUE),
    child_age_median = median(child_age_years, na.rm = TRUE), 
    child_age_sd = sd(child_age_years, na.rm = TRUE),
    min_child_age = min(child_age_years, na.rm = TRUE), 
    n = n(), 
    .groups = "drop")

# visual 
ggplot(df_bite, aes(
  x = factor(bite_severity_clean),
  y = child_age_years)) + 
  geom_boxplot() + geom_jitter(
    width = 0.15,
    alpha = 0.7,
    size = 2) + 
  labs(
    title = "Child Age Across Bite Severity",
    x = "Bite Severity",
    y = "Child Age (Years)") + 
  theme_minimal()
```

## child age x warning signals
```{r}
# summary table
df_bite %>%
  group_by(dog_warning_any) %>%
  summarise(
    mean_child_age = mean(child_age_years, na.rm = TRUE),
    median_child_age = median(child_age_years, na.rm = TRUE),
    sd_child_age = sd(child_age_years, na.rm = TRUE),
    min_child_age = min(child_age_years, na.rm = TRUE),
    max_child_age = max(child_age_years, na.rm = TRUE),
    n = n(),
    .groups = "drop")

# visualization
ggplot(df_bite,
       aes(
         x = dog_warning_any,
         y = child_age_years)) +
  geom_boxplot() +
  geom_jitter(
    width = 0.15,
    alpha = 0.7,
    size = 2) +
  scale_x_discrete(labels = c(
    "FALSE" = "No warning signal",
    "TRUE" = "Warning signal")) +
  labs(
    title = "Child Age by Warning Signal Presence",
    x = "Warning Signal Presence",
    y = "Child Age (Years)") +
  theme_minimal()
```

## dog age x severity
```{r}
# summary table
df_bite %>%
  group_by(bite_severity_clean) %>%summarise(
    mean_dog_age = mean(dog_age_years, na.rm = TRUE),
    median_dog_age = median(dog_age_years, na.rm = TRUE),
    sd_dog_age = sd(dog_age_years, na.rm = TRUE),
    min_dog_age = min(dog_age_years, na.rm = TRUE),
    max_dog_age = max(dog_age_years, na.rm = TRUE),
    n = n(),
    .groups = "drop")

# visualization
ggplot(df_bite, aes(
         x = factor(bite_severity_clean),
         y = dog_age_years)) +
  geom_boxplot() +
  geom_jitter(
    width = 0.15,
    alpha = 0.7,
    size = 2) +
  labs(
    title = "Dog Age Across Bite Severities",
    x = "Bite Severity",
    y = "Dog Age (Years)") +
  theme_minimal()
```
```{r dog-age-x-warning}
## dog age x warning signs 
# summary table
df_bite %>%
  group_by(dog_warning_any) %>% summarise(
    mean_dog_age = mean(dog_age_years, na.rm = TRUE),
    median_dog_age = median(dog_age_years, na.rm = TRUE),
    sd_dog_age = sd(dog_age_years, na.rm = TRUE),
    min_dog_age = min(dog_age_years, na.rm = TRUE),
    max_dog_age = max(dog_age_years, na.rm = TRUE),
    n = n(),
    .groups = "drop")

# visualization
ggplot(df_bite, aes(
         x = dog_warning_any,
         y = dog_age_years)) +
  geom_boxplot() +
  geom_jitter(
    width = 0.15,
    alpha = 0.7,
    size = 2) +
  scale_x_discrete(labels = c(
    "FALSE" = "No warning signal",
    "TRUE" = "Warning signal")) +
  labs(
    title = "Dog Age by Warning Signal Presence",
    x = "Warning Signal Presence",
    y = "Dog Age (Years)") +
  theme_minimal()
```


```{r dog-gender-x-severity}
## dog gender x severity
# counts
df_bite %>% count(dog_sex_std, bite_severity_clean)

# visualization
ggplot(df_bite, aes(
         x = dog_sex_std,
         fill = factor(bite_severity_clean))) +
  geom_bar(position = "stack") +
  scale_x_discrete(labels = c(
    Female = "Female",
    Male = "Male")) +
  labs(
    title = "Bite Severity by Dog Sex",
    x = "Dog Sex",
    y = "Count",
    fill = "Severity") +
  theme_minimal()
```

```{r dog-gender-signal}
## dog gender x signals
# counts
df_bite %>% count(dog_sex_std, dog_warning_any)

# proportional visualization
ggplot(df_bite, aes(
         x = dog_sex_std,
         fill = dog_warning_any)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_discrete(labels = c(
    "FALSE" = "No warning signal",
    "TRUE" = "Warning signal")) +
  scale_x_discrete(labels = c(
    Female = "Female",
    Male = "Male")) +
  labs(
    title = "Warning Signal Presence by Dog Sex",
    x = "Dog Sex",
    y = "Proportion of Cases",
    fill = "Warning Signal") +
  theme_minimal()
```

## breed x warning signal

## breed type x severity

### MISSING data on what age the dog was when they started living in their current home

